---
title: "Figure 3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

In this vignette, we'll take the tcells cluster centers and project them onto two CITE-seq datasets: a dataset of 30,000 human bone marrow cells available through `SeuratData::bmcite`, and then we'll use a larger dataset of 162,000 PBMCs with a panel of 228 antibodies.

Finally, we'll also use a dataset of 40,000 cells from the Human Immune Cell atlas also available through `SeuratData::hcabm40k`.

Both of these datasets are described and explored in the Seurat vignette on multimodal analysis, which you can find [https://satijalab.org/seurat/articles/multimodal_reference_mapping.html](here).

## Bmcite analysis

1. Project bmcite using a binary L0-truncated projection.

```{r}
# install SeuratData
# library(devtools)
# devtools::install_github('satijalab/seurat-data')
library(seuratData)
library(Seurat)

# download bmcite data
InstallData("bmcite")
bmcite <- LoadData("bmcite")

bmcite <- RunUMAP(bmcite, dims = 1:20, reduction = "spca")
bm_ann <- bmcite@reductions$umap@cell.embeddings
bm_ann <- data.frame(cbind(bm_ann, bmcite$celltype.l1, bmcite$celltype.l2))
colnames(bm_ann) <- c("umap1", "umap2", "celltype1", "celltype2")

A_adt <- bmcite@assays$ADT@data
A_rna <- bmcite@assays$RNA@data

# get tcell annotated centers
data(tcells)

# get intersection between tcells and w
genes_in_common <- intersect(rownames(A_rna), rownames(tcells))
A_rna <- A_rna[genes_in_common,]
w <- tcells[genes_in_common,]

# tlearn projection
h <- tlearn(A_rna, w, L0 = 1)
hplot <- mheatmap(h[,sample(1:ncol(h), 5000)], Colv = TRUE, Rowv = TRUE)
hplot <- mheatmap(h_cos[,sample(1:ncol(h), 1000)], Colv = TRUE, Rowv = TRUE, show.rownames = TRUE, show.colnames = FALSE)

# have a look at the values that are being projected
maxvals <- apply(h, 2, max)

# find the best binary value


h_cos <- cosine(A_rna, w)

```

## Download data

Load the t-cells data, available through `tlearn`:

```{r}
# install SeuratData
# library(devtools)
# devtools::install_github('satijalab/seurat-data')
library(seuratData)

# install SeuratDisk
# devtools::install_github("mojaveazure/seurat-disk")
library(SeuratDisk)
library(Seurat)

# download bmcite data
InstallData("bmcite")
bmcite <- LoadData("bmcite")

# download hcabm40k data
InstallData("hcabm40k")
hcabm40k <- LoadData("hcabm40k")

# download files
# https://atlas.fredhutch.org/data/nygc/multimodal/pbmc_multimodal.h5seurat
# move to working directory as "pbmc_multimodal.h5seurat"

  assay SCT -> reductions -> umap
  SCT -> data

  ADT -> data

pbmc_sct <- LoadH5Seurat("data/pbmc_multimodal.h5Seurat", assays = list(SCT = c("data")))

pbmc_sct <- LoadH5Seurat("data/pbmc_multimodal.h5Seurat", assays = list(SCT = c("data")))

```

```{r}
data(tcells)
```